Making Sequential Programs Deterministic

### 2019-03-23 design.tex
design: climb to top

The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overriden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overriden method (\<@B>).

@@

\subsection{Dataflow analysis and type refinement}\label{dataflow}
The checker framework performs automatic type inference for local types within method bodies.
It also performs dataflow analysis and refines the types of these locals, thereby reducing the annotation burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       //  No error here.
}
\end{verbatim}
The checker framework performs dataflow analysis within method bodies and refines the types of local variables.
In the example above, even though the local variable \codeid{y} was declared to be \codeid{@NonDet}, it gets
type refined to \codeid{@Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{@Det} variable \codeid{z} is sound. In the follwing example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
the checker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

@@

\begin{itemize}
	\item Local variables are defaulted to top because type refinement is applied to local variables. If a local variable starts as the 
	\textit{top} type, then the Checker Framework refines it to the best (most specific) possible type based on assignments to it. As a 
	result, a programmer rarely writes an explicit annotation on any of those locations.
	Variables defaulted to top include \textit{local variables}, resource variables in the \textit{try-with-resources} construct, variables 
	in \textit{for} statements, and \textit{catch} arguments (known as exception parameters in the Java Language Specification). 
	\textit{Exception parameters} need to have the \textit{top} type because exceptions of arbitrary qualified types can be thrown 
	and the Checker Framework does not provide runtime checks.
	\item \textit{Cast} and \textit{instanceof} types are given the same type as their argument expression. This has the same effect as 
	if they were given the \textit{top} type and then flow-sensitively refined to the type of their argument.
	\item Implicit upper bounds are defaulted to top to allow them to be instantiated in any way. If a user declared class \codeid{C<T> 
		{ ... }}, then the Checker Framework assumes that the user intended to allow any instantiation of the class, and the declaration
	is interpreted as class \codeid{C<T extends @NonDet Object> { ... }} rather than as class \codeid{C<T extends @Det Object> { ... }}. The latter would forbid instantiations such as \codeid{C<@Det String>}, or would require rewriting of code. On the other hand, if a user writes 
	an explicit bound such as class \codeid{C<T extends D> { ... }}, then the user intends some restriction on instantiation and can write a 
	qualifier on the upper bound as desired.
	This rule means that the upper bound of class \codeid{C<T>} is defaulted differently than the upper bound of class \codeid{C<T extends Object>}. 
	This may seem confusing, but it is the least bad option. The more confusing alternative would be for \codeid{Object} to be defaulted 
	differently in class \codeid{C<T extends Object>} and in an instantiation \codeid{C<Object>}, and for the upper bounds to be defaulted differently 
	in class \codeid{C<T extends Object>} and class \codeid{C<T extends Date>}.
	\item \textit{Implicit lower bounds} are defaulted to the \textit{bottom} type, again to allow maximal instantiation. Note that Java does not allow a programmer to express both the upper and lower bounds of a type, but the Checker Framework allows the programmer to specify either or both.
\end{itemize}
With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<@NonDet>.

\begin{itemize}
	\item So we default locals collections to @NonDet<@NonDet>. 
	\item This is sound but could produce a lot of FPs. Example assignment: @NonDet<@NonDet> collection = collection1.
	Type of collection1 is @Det<@Det> and it is a method argument.
	\item To avoid this, we require users to explicitly annotate local collections.
	\item Similarly for local arrays.
\end{itemize}

>>>>

The overriding method is invalid for two reasons: type of the parameter (\<@B>) is not a subtype of the parameter type in the overridden method (\<@C>) and the return type (\<@C>) is not a supertype of the return type of the overridden method (\<@B>).

@@

\subsection{Dataflow analysis and type refinement}\label{dataflow}
The checker framework performs automatic type inference for local types within method bodies.
It also performs dataflow analysis and refines the types of these locals, thereby reducing the annotation burden 
on the programmer. Consider the example below:
\begin{verbatim}
void foo(@Det int x) {
    @NonDet int y = x;    // Type of y gets refined to @Det.
    @Det int z = y;       //  No error here.
}
\end{verbatim}
The checker framework performs dataflow analysis within method bodies and refines the types of local variables.
In the example above, even though the local variable \codeid{y} was declared to be \codeid{@NonDet}, it gets
type refined to \codeid{@Det} after the first assignment statement. As the result, the assignment of \codeid{y}
to a \codeid{@Det} variable \codeid{z} is sound. In the following example,
\begin{verbatim}
void foo(@NonDet Object x, @Det Object y) {
    x = y;                // No type refinement for x.
    @Det Object z = x;    // Error.
}
\end{verbatim}
the checker reports a warning at the second assignment statement of \codeid{foo}. Type refinement only applies to locals. Refining non-locals can result in side effects making the analysis unsound.

@@

\begin{itemize}
	\item Local variables are defaulted to top because type refinement is applied to local variables. If a local variable starts as the 
	\textit{top} type, then the Checker Framework refines it to the best (most specific) possible type based on assignments to it. As a 
	result, a programmer rarely writes an explicit annotation on any of those locations.
	Variables defaulted to top include \textit{local variables}, resource variables in the \textit{try-with-resources} construct, variables 
	in \textit{for} statements, and \textit{catch} arguments (known as exception parameters in the Java Language Specification). 
	\textit{Exception parameters} need to have the \textit{top} type because exceptions of arbitrary qualified types can be thrown 
	and the Checker Framework does not provide runtime checks.
	\item \textit{Cast} and \textit{instanceof} types are given the same type as their argument expression. This has the same effect as 
	if they were given the \textit{top} type and then flow-sensitively refined to the type of their argument.
	\item Implicit upper bounds are defaulted to top to allow them to be instantiated in any way. If a user declared class \codeid{C<T> 
		{ ... }}, then the Checker Framework assumes that the user intended to allow any instantiation of the class, and the declaration
	is interpreted as class \codeid{C<T extends @NonDet Object> { ... }} rather than as class \codeid{C<T extends @Det Object> { ... }}. The latter would forbid instantiations such as \codeid{C<@Det String>}, or would require rewriting of code. On the other hand, if a user writes 
	an explicit bound such as class \codeid{C<T extends D> { ... }}, then the user intends some restriction on instantiation and can write a 
	qualifier on the upper bound as desired.
	This rule means that the upper bound of class \codeid{C<T>} is defaulted differently than the upper bound of class \codeid{C<T extends Object>}. 
	This may seem confusing, but it is the least bad option. The more confusing alternative would be for \codeid{Object} to be defaulted 
	differently in class \codeid{C<T extends Object>} and in an instantiation \codeid{C<Object>}, and for the upper bounds to be defaulted differently 
	in class \codeid{C<T extends Object>} and class \codeid{C<T extends Date>}.
	\item \textit{Implicit lower bounds} are defaulted to the \textit{bottom} type, again to allow maximal instantiation. Note that Java does not allow a programmer to express both the upper and lower bounds of a type, but the Checker Framework allows the programmer to specify either or both.
\end{itemize}
With the above rules, the checker framework will automatically annotate a local collection with type \<@NonDet> and its
type arguments will get the type \<@Det>. For example, a locally declared \codeid{List} of \codeid{Strings} will be automatically annotated
as \codeid{@NonDet List<@Det String>}. Recall from~\ref{collection-rules} that the determinism checker does not allow this type on
any collection. While this is still sound, it will result in the determinism checker reporting a lot of errors because every locally declared 
collection gets annotated with this invalid type. To avoid this situation, we make an exception to the defaulting rules for local collections
and annotate their type parameters also with \<@NonDet>. So a local list of Strings gets the default annotation of \codeid{@NonDet List<@NonDet String>}. While this defaulting eliminates the possibility of automatically annotating locals collections with an invalid
type, it would could in a high number of false positives. In the following code snippet,
\begin{verbatim}
void test(@Det List<@Det String> argList) {
            List<String> localList = argList;
}    
\end{verbatim}
the checker would report an error at the assignment statement since the inferred type for \codeid{localList} is 
\codeid{@NonDet List<@NonDet> String} and since collections types are invariant.
We therefore recommend that programmers explicitly annotate
all local variables that are collections to reduce false positives. The same reasoning and recommendation applies to arrays.
